### 2023-07-13 content/submission/plos_revision1/answer_to_reviewers.md

> How was the sanity check performed (as mentioned in the second paragraph of the subsection "Preparation for Analysis")?

The mentioned 'sanity check' was performed manually, e.g. by looking at all the values (mostly the reconstruction parameters), to exclude any operator error.
If errors were spotted at this stage, the reconstructions were deleted and re-done with the correct parameters.
Sometimes errors or problems were only spotted at a later stage, when either the pharyngeal or oral jaws were extracted, or when we noticed that some part of the the whole skull were out of the field of view of the tomographic scan prior to performing the PCA analysis mentioned later in the manuscript.
This means that specifying this as 'sanity check' is more of a glorified wording of telling the readers that we manually looked at the values of each performed tomographic scan to exlude and catch operator errors.
We reworded the sentence to hopefully make this more clear.

> In the same paragraph, the authors mentioned that each tomographic dataset contains around 2700 slices, exceeding the available RAM size on an average high-end workstation.
> Could you clarify if this limitation affected the analysis and its implications?

We wrote that *the total size of the acquired data exceeds the RAM* of any workstation (e.g. around 1.5 TB).
*One* or more single datasets would fit into RAM very well, It is not possible to load *all* data concurrently for analysis.
The use of *Dask* and more specifically [*dask-image*](@https://image.dask.org/) for loading only the currently used data for each specimen from the HD (or remote storage) into RAM on demand made working with the *all* the data feasible.
We have updated the sentence to explain this.

### 2023-07-13 manuscript.tex
Whitespace changes only [ci skip]
This build is based on
https://github.com/habi/EAWAG-manuscript/commit/a15f6a93f7f5b262f005c909a8fdcbdf8220da53.

This commit was created by the following CI build and job:
https://github.com/habi/EAWAG-manuscript/commit/a15f6a93f7f5b262f005c909a8fdcbdf8220da53/checks
https://github.com/habi/EAWAG-manuscript/actions/runs/5542383142

\emph{A preprint of this manuscript is available at bioRχiv with the \href{https://doi.org/10.1101/2023.03.30.534917}{doi:10.1101/2023.03.30.534917}.}
\emph{We submitted \href{https://github.com/habi/EAWAG-manuscript/releases/tag/v1.0}{\texttt{v1.0}} (\href{https://habi.github.io/EAWAG-manuscript/v/e1e2ef76a476174a4115937d77457037ddec95df/}{online version}) of the manuscript to \href{https://journals.plos.org/plosone/}{PLOS ONE}.}
\emph{The version you see here \href{https://github.com/habi/EAWAG-manuscript/compare/v1.0...05e0c4de935ffd552c20ae7ff34db6d16ce5ca6d}{has been updated} since submission.}

This manuscript
(\href{https://habi.github.io/EAWAG-manuscript/v/05e0c4de935ffd552c20ae7ff34db6d16ce5ca6d/}{permalink})
was automatically generated
from \href{https://github.com/habi/EAWAG-manuscript/tree/05e0c4de935ffd552c20ae7ff34db6d16ce5ca6d}{habi/EAWAG-manuscript@05e0c4d}
on July 11, 2023.

@@

\hypertarget{history}{%
\subsubsection{History}\label{history}}

@@

X-ray microtomography is a valuable tool to gain insights into the inner structure of highly diverse samples, namely for specimens related to research done in the biomedical sciences.
Microtomographic imaging has been employed as a method of choice to nondestructively assess the morphology of different fish species, large and small {[}\protect\hyperlink{ref-19h9vxsYG}{6}{]}.
For a small overview of analyses which are possible with X-ray microtomographic imaging in relation to fish biology and morphology, see prior work of the authors of this manuscript {[}\protect\hyperlink{ref-19h9vxsYG}{6}--\protect\hyperlink{ref-BDrCSu8p}{8}{]} as well as other studies {[}\protect\hyperlink{ref-3rxGCEvJ}{9}{]}.

Depending on the structures of interest, biomedical samples are often only tomographically scanned after being stained with a contrast agent, most often employing contrast agents containing heavy metals.
Since the structures of interest for the two studies we touch upon in this manuscript (cichlid teeth and skulls) display large enough contrast to the surrounding tissue we did not stain our samples prior to the tomographic imaging presented here.

@@

The fish were sorted into `bins' based on their physical size.
We used a custom-made sample-holder to scan each of the fish in our machines.
It was 3D-printed on a Form 2 desktop stereolithography printer (Formlabs, Somerville, Massachusetts, USA), the file for printing the holder is available online {[}\protect\hyperlink{ref-VCZPOv2f}{10}{]} as part of a library of sample holders for tomographic scanning of biomedical samples {[}\protect\hyperlink{ref-115PPSuQp}{11}{]}.
The original \emph{OpenSCAD} {[}\protect\hyperlink{ref-wsjdcTeC}{12}{]} file {[}\protect\hyperlink{ref-f7OzrzUv}{13}{]} is parametrized to effortlessly generate a file for 3D-printing sample holders to accommodate the varying width, height and length classes of the fish.

@@

A small script {[}\protect\hyperlink{ref-ZjgrPHTa}{14}{]} was used to generate redundant (archival) copies of the raw projection images and copy all the files to a shared network drive on the `Research Storage' infrastructure of the University of Bern, enabling collaboration on the data by all authorized persons at the same time.
This automated archival and copying process made it possible to delete the raw projection images on the production system shortly after acquisition, freeing resources and easing data handling.
A subset of the data, namely the log files and reconstructed image stacks were kept on the production system for working with it (as described in \protect\hyperlink{preparation-for-analysis}{Preparation for analysis} below).

@@

We wrote a set of \emph{Jupyter} {[}\protect\hyperlink{ref-pQ6Wbz73}{15}{]} notebooks with \emph{Python} code to work with the acquired images and to extract the desired data.
The notebooks were written at the start of the project, to be able to process new scans as soon as they were reconstructed.
Re-runs of the notebook added newly scanned and reconstructed data to the analysis, facilitating an efficient quality check of the scans and batched processing of the data.

All notebooks written for this work are available online {[}\protect\hyperlink{ref-1HteOscVd}{16}{]} and are extensively commented.
They are split into different task performed for this study and mentioned in their respective sections below.

@@

At first, \emph{all} log files of \emph{all} the scan present in the processed folder were read into a \emph{pandas} {[}\protect\hyperlink{ref-19qQOHlFN}{17},\protect\hyperlink{ref-U5AcoDOX}{18}{]} dataframe.
This already enabled us to extract the specimen name and scan.
Since we performed multiple scans for each specimen, i.e.~a low resolution scan with large field of view for the whole head and one or two scans in high resolution focusing on the region of the oral and pharyngeal jaws it was necessary to be able to efficiently find the scan to be looked at in detail.
From the log files we extracted the scanning and reconstruction parameters of each performed scan for double-checking them and excluding (or necessarily correcting) any operator-error.
All relevant values for each scan were also saved into the aforementioned ataframe.
This allows for an easy collation of all the relevant values into a table (as shown in the \protect\hyperlink{supplementary-materials}{Supplementary Materials}) at the end of each run of the notebook.

After displaying several parameters of both data acquisition and reconstruction for ruling out any operator error we used \emph{Dask} {[}\protect\hyperlink{ref-hj1CnyWB}{19}{]} to efficiently access the tomographic data on disk (in the end amounting to a total of nearly a million single images).
On average, each of the tomographic datasets contains around 2700 slices, so the total size of the acquired data exceeds the RAM size available on an average high-end workstation.

@@

To extract the oral jaw (OJ) and pharyngeal jaw (PJ) of the fish, we used \href{https://www.slicer.org}{\emph{3D Slicer}} (Version 4.11.20210226) {[}\protect\hyperlink{ref-OsCpHsJH}{20}{]} extended with the \emph{SlicerMorph} tools {[}\protect\hyperlink{ref-MwdN6kPV}{21}{]} which aim to help biologists work with 3D specimen data.
The reconstructed image stacks were loaded into \href{https://www.slicer.org/wiki/Documentation/Labs/ImageStacks}{\emph{ImageStacks}}, depending on their size we reduced the image resolution (i.e.~downscaled the images) for this first step.
The three-dimensional volume was rendered via \href{https://slicer.readthedocs.io/en/latest/user_guide/modules/volumerendering.html}{\emph{VTK GPU Ray Casting}}.
A custom-made volume property was used as an input to view the scans.
Using toggles in the volume rendering, we defined regions of interest (ROIs) for both the OJs and PJs in each specimen.
These ROIs were then extracted in their native resolution from the original dataset for further processing.
Using the gray value thresholding function in \emph{3D Slicer}'s \href{https://slicer.readthedocs.io/en/latest/user_guide/modules/segmenteditor.html}{\emph{Segment Editor}}, the teeth in both the oral and pharyngeal jaws were extracted.
We used the \emph{Scissor} and \emph{Island} tools of the \emph{Segment Editor} to isolate single regions.

Processed regions of interest were exported as \texttt{Nrrd} {[}\protect\hyperlink{ref-rn6XbmFW}{22}{]} files.
The three-dimensional visualizations of all regions of interest for each specimen were compiled into overview images (see Figure \ref{fig:104016} for an example from the compilation document).
In total, we compiled such overview images for 133 specimens with full head morphology, oral jaw and lower pharyngeal jaw profiles.

\begin{figure}
\hypertarget{fig:104016}{%
\centering
\includegraphics{images/104016.png}
\caption{Overview of data from sample 104016,~\emph{Enterochromis I cinctus}, from `Station E' in the transect of Mwanza Bay at the southern edge of Lake Victoria.
A map showing the detailed location of `Station E' is shown in Fig. 1 of {[}\protect\hyperlink{ref-JtHcU2O2}{23}{]}.
Panel A: Three-dimensional visualization of the head scan.
Panel B: Three-dimensional visualization of the oral jaw scan.
Panel C: Photograph of the specimen.
Panels D and E: Three-dimensional visualization of the pharyngeal jaw, dorsoventral and lateral view, respectively.}\label{fig:104016}
}
\end{figure}

@@

Current studies are using 3D geometric morphometrics to compare the morphological shape of these scanned cichlids using statistical analysis.
We used a homologous landmark scheme across one-half of the skull for higher density of shape information {[}\protect\hyperlink{ref-BDrCSu8p}{8},\protect\hyperlink{ref-15FafMZm5}{24}{]}, and landmarks were placed on each specimen using \emph{3D Slicer}.
To examine differences in shape across the species sampled, we performed a Generalized Procrustes Superimposition on the landmark data to remove the effects of location, size, and rotation from the analysis using the \emph{geomorph} package in \emph{R} (Version 4.2.1) with \emph{RStudio} (Version 2022.07.2+576) {[}\protect\hyperlink{ref-VPUOQA2P}{25}--\protect\hyperlink{ref-EwNL0C0Y}{27},\protect\hyperlink{ref-whEyWyQh}{28},\protect\hyperlink{ref-QfVloFVu}{29}{]}.
This process brings all specimens to a common origin, scales the landmarks to a unit centroid size, and rotates specimens to reduce distances between landmarks.
A principal components analysis was then performed in \emph{geomorph} on the superimposed landmark data to visualize the major axes of shape change across sampled species.
We then used phylogenetic information to identify instances of repeated evolution of trophic adaptations in these cichlids.

@@

Otoliths are structures made up of mostly calcium carbonate located in the head of fishes.
Due to their composition, they are easily distinguished in the X-ray images we acquired.
We devised an image processing method to automatically and robustly detect the location of the otoliths in the tomographic scans of the skulls of the cichlids.
Based on this detected location we were then able to efficiently extract a cropped region of interest from the original data which includes only the region of the otoliths.
The whole method is implemented in its own \href{https://github.com/habi/EAWAG/blob/master/ExtractOtoliths.ipynb}{\emph{Jupyter} notebook} (part of the aforementioned analysis repository {[}\protect\hyperlink{ref-1HteOscVd}{16}{]}).

@@

By detecting the largest components in the cropped copies of the datasets we can easily extract and visualize the otoliths in 3D, as shown in Figure \ref{fig:otolith3d}.
The extracted otoliths are thus prepared for further analysis and visualization.
The simple three-dimensional visualization is integrated in the aforementioned \emph{Jupyter} notebook through an integrated visualization library {[}\protect\hyperlink{ref-a9ZL41Of}{30}{]} and is also shown in the \protect\hyperlink{supplementary-materials}{Supplementary Materials}.

@@

The notebook for extracting the otoliths can be run in your browser without installing any additional software (via \emph{Binder} {[}\protect\hyperlink{ref-Q20Bxdsr}{31}{]}).
To do this, one starts the notebook by \href{https://mybinder.org/v2/gh/habi/eawag/HEAD?labpath=ExtractOtoliths.ipynb}{clicking a single button} in the \href{https://github.com/habi/EAWAG/blob/master/README.md}{README file} of the \href{https://github.com/habi/EAWAG}{project repository} {[}\protect\hyperlink{ref-1HteOscVd}{16}{]}.
This starts a computing environment in the cloud, downloads the tomographic data we acquired of \emph{one} specimen, and performs both the otolith extraction and visualization in your browser.

@@

The presented method offers an insight and algorithm on how to perform tomographic scans, preview and analyze micro-computer tomographic datasets of a large collection of fish.
The workflow is relying only on free and open-source software and can thus be used and verified independently by any interested reader.
All the \emph{Jupyter} notebook described herein is also freely available online {[}\protect\hyperlink{ref-1HteOscVd}{16}{]}.

@@

We thank Salome Mwaiko for taking care of the fish collection at Eawag and Mark Charran for helping to find suitable specimens to represent each species.
We are grateful to the \href{https//www.mic.unibe.ch/}{Microscopy Imaging Center} of the University of Bern for the infrastructural support.
We thank the \texttt{manubot} project {[}\protect\hyperlink{ref-YuJbg3zO}{32}{]} for helping us write this manuscript collaboratively.

@@

\leavevmode\vadjust pre{\hypertarget{ref-8Pmc2mI8}{}}%
\CSLLeftMargin{7. }
\CSLRightInline{Messerli M, Aaldijk D, Haberthür D, Röss H, García-Poyatos C, Sande-Melón M, et al. Adaptation mechanism of the adult zebrafish respiratory organ to endurance training. Shelden EA, editor. PLoS ONE. 2020;15: e0228333. doi:\href{https://doi.org/10.1371/journal.pone.0228333}{10.1371/journal.pone.0228333}}

\leavevmode\vadjust pre{\hypertarget{ref-BDrCSu8p}{}}%
\CSLLeftMargin{8. }
\CSLRightInline{Ford KL, Peterson R, Bernt M, Albert JS. Convergence is Only Skin Deep: Craniofacial Evolution in Electric Fishes from South America and Africa (Apteronotidae and Mormyridae). Integrative Organismal Biology. 2022;4. doi:\href{https://doi.org/10.1093/iob/obac022}{10.1093/iob/obac022}}

\leavevmode\vadjust pre{\hypertarget{ref-3rxGCEvJ}{}}%
\CSLLeftMargin{9. }
\CSLRightInline{Available: \url{https://osf.io/ecmz4}}

\leavevmode\vadjust pre{\hypertarget{ref-VCZPOv2f}{}}%
\CSLLeftMargin{10. }
\CSLRightInline{Available: \url{https://github.com/TomoGraphics/Hol3Drs/blob/master/STL/EAWAG.Fish.stl}}

\leavevmode\vadjust pre{\hypertarget{ref-115PPSuQp}{}}%
\CSLLeftMargin{11. }
\CSLRightInline{Haberthür D. TomoGraphics/Hol3Drs: A release. Zenodo; 2019. doi:\href{https://doi.org/10.5281/zenodo.2587555}{10.5281/zenodo.2587555}}

\leavevmode\vadjust pre{\hypertarget{ref-wsjdcTeC}{}}%
\CSLLeftMargin{12. }
\CSLRightInline{OpenSCAD. {[}cited 11 Jul 2023{]}. Available: \url{https://openscad.org}}

\leavevmode\vadjust pre{\hypertarget{ref-f7OzrzUv}{}}%
\CSLLeftMargin{13. }
\CSLRightInline{Available: \url{https://github.com/TomoGraphics/Hol3Drs/blob/master/EAWAG.Fish.scad}}

\leavevmode\vadjust pre{\hypertarget{ref-ZjgrPHTa}{}}%
\CSLLeftMargin{14. }
\CSLRightInline{Available: \url{https://github.com/habi/EAWAG/blob/master/rsync-fishes.sh}}

\leavevmode\vadjust pre{\hypertarget{ref-pQ6Wbz73}{}}%
\CSLLeftMargin{15. }
\CSLRightInline{Kluyver T, Ragan-Kelley B, Pérez F, Granger B, Bussonnier M, Frederic J, et al. Jupyter Notebooks -- a publishing format for reproducible computational workflows. In: Loizides F, Scmidt B, editors. IOS Press; 2016. pp. 87--90. doi:\href{https://doi.org/10.3233/978-1-61499-649-1-87}{10.3233/978-1-61499-649-1-87}}

\leavevmode\vadjust pre{\hypertarget{ref-1HteOscVd}{}}%
\CSLLeftMargin{16. }
\CSLRightInline{Haberthür D. habi/EAWAG: Jupyter notebooks for preparing tomographic scans of a collection of cichlids for analysis. Zenodo; 2022. doi:\href{https://doi.org/10.5281/zenodo.6798632}{10.5281/zenodo.6798632}}

\leavevmode\vadjust pre{\hypertarget{ref-19qQOHlFN}{}}%
\CSLLeftMargin{17. }
\CSLRightInline{Team TPD. pandas-dev/pandas: Pandas. Zenodo; 2022. doi:\href{https://doi.org/10.5281/zenodo.7093122}{10.5281/zenodo.7093122}}

\leavevmode\vadjust pre{\hypertarget{ref-U5AcoDOX}{}}%
\CSLLeftMargin{18. }
\CSLRightInline{McKinney W. Data Structures for Statistical Computing in Python. Proceedings of the Python in Science Conference. SciPy; 2010. doi:\href{https://doi.org/10.25080/majora-92bf1922-00a}{10.25080/majora-92bf1922-00a}}

\leavevmode\vadjust pre{\hypertarget{ref-hj1CnyWB}{}}%
\CSLLeftMargin{19. }
\CSLRightInline{Dask Development Team. Dask: Library for dynamic task scheduling. 2016. Available: \url{https://dask.org}}

\leavevmode\vadjust pre{\hypertarget{ref-OsCpHsJH}{}}%
\CSLLeftMargin{20. }
\CSLRightInline{Fedorov A, Beichel R, Kalpathy-Cramer J, Finet J, Fillion-Robin J-C, Pujol S, et al. 3D Slicer as an image computing platform for the Quantitative Imaging Network. Magnetic Resonance Imaging. 2012;30: 1323--1341. doi:\href{https://doi.org/10.1016/j.mri.2012.05.001}{10.1016/j.mri.2012.05.001}}

>>>>

\emph{A preprint of this manuscript is available at bioRχiv with the \href{https://doi.org/10.1101/2023.03.30.534917}{doi:10.1101/2023.03.30.534917}.}
\emph{We submitted \href{https://github.com/habi/EAWAG-manuscript/releases/tag/v1.0}{\texttt{v1.0}} (\href{https://habi.github.io/EAWAG-manuscript/v/e1e2ef76a476174a4115937d77457037ddec95df/}{online version}) of the manuscript to \href{https://journals.plos.org/plosone/}{PLOS ONE}.}
\emph{The version you see here \href{https://github.com/habi/EAWAG-manuscript/compare/v1.0...a15f6a93f7f5b262f005c909a8fdcbdf8220da53}{has been updated} since submission.}

This manuscript
(\href{https://habi.github.io/EAWAG-manuscript/v/a15f6a93f7f5b262f005c909a8fdcbdf8220da53/}{permalink})
was automatically generated
from \href{https://github.com/habi/EAWAG-manuscript/tree/a15f6a93f7f5b262f005c909a8fdcbdf8220da53}{habi/EAWAG-manuscript@a15f6a9}
on July 13, 2023.

@@

\hypertarget{history-and-rationale}{%
\subsubsection{History and rationale}\label{history-and-rationale}}

@@

X-ray microtomography is a valuable tool to gain insights into the inner structure of highly diverse samples, namely for specimens related to the biomedical sciences {[}\protect\hyperlink{ref-IKVBgaCE}{7}{]}.
Microtomographic imaging has been employed as a method of choice to nondestructively assess the morphology of different fish species, large and small {[}\protect\hyperlink{ref-19h9vxsYG}{6}{]}.
For a small overview of analyses which are possible with X-ray microtomographic imaging in relation to fish biology and morphology, see prior work of the authors of this manuscript {[}\protect\hyperlink{ref-19h9vxsYG}{6},\protect\hyperlink{ref-8Pmc2mI8}{8},\protect\hyperlink{ref-BDrCSu8p}{9}{]} as well as other studies {[}\protect\hyperlink{ref-3rxGCEvJ}{10}{]}.

Depending on the structures of interest, biomedical samples are often only tomographically scanned after being stained with a contrast agent {[}\protect\hyperlink{ref-pN1S9E5p}{11}{]}, most often employing contrast agents containing heavy metals {[}\protect\hyperlink{ref-kkBa4qNV}{12}{]}.
Due to their inherent contrast difference to the surrounding tissue, the structures of interest we touch upon in this manuscript (teeth and bones, i.e.~jaws and skull) are well visualized in native samples.
Enhancing contrast differences by staining prior to tomographic imaging was thus not necessary to for the samples presented herein.

@@

The fish were sorted into `bins' based on their physical size.
We used a custom-made sample-holder to scan each of the fish in our machines.
It was 3D-printed on a Form 2 desktop stereolithography printer (Formlabs, Somerville, Massachusetts, USA), the file for printing the holder is available online {[}\protect\hyperlink{ref-VCZPOv2f}{13}{]} as part of a library of sample holders for tomographic scanning of biomedical samples {[}\protect\hyperlink{ref-115PPSuQp}{14}{]}.
The original \emph{OpenSCAD} {[}\protect\hyperlink{ref-wsjdcTeC}{15}{]} file {[}\protect\hyperlink{ref-f7OzrzUv}{16}{]} is parametrized to effortlessly generate a file for 3D-printing sample holders to accommodate the varying width, height and length classes of the fish.

@@

A small script {[}\protect\hyperlink{ref-ZjgrPHTa}{17}{]} was used to generate redundant (archival) copies of the raw projection images and copy all the files to a shared network drive on the `Research Storage' infrastructure of the University of Bern, enabling collaboration on the data by all authorized persons at the same time.
This automated archival and copying process made it possible to delete the raw projection images on the production system shortly after acquisition, freeing resources and easing data handling.
A subset of the data, namely the log files and reconstructed image stacks were kept on the production system for working with it (as described in \protect\hyperlink{preparation-for-analysis}{Preparation for analysis} below).

@@

We wrote a set of \emph{Jupyter} {[}\protect\hyperlink{ref-pQ6Wbz73}{18}{]} notebooks with \emph{Python} code to work with the acquired images and to extract the desired data.
The notebooks were written at the start of the project, to be able to process new scans as soon as they were reconstructed.
Re-runs of the notebook added newly scanned and reconstructed data to the analysis, facilitating an efficient quality check of the scans and batched processing of the data.

All notebooks written for this work are available online {[}\protect\hyperlink{ref-1HteOscVd}{19}{]} and are extensively commented.
The modularized and interactive format chosen for these notebooks makes them easy to adapt for any other study dealing with tomographic data.
The first author of this study is happy to help other scientists to adapt the notebooks to their use-case.
They notebooks are split into different task performed for this study and are specifically mentioned in the respective sections below.

@@

At first, \emph{all} log files of \emph{all} the scan present in the processed folder were read into a \emph{pandas} {[}\protect\hyperlink{ref-19qQOHlFN}{20},\protect\hyperlink{ref-U5AcoDOX}{21}{]} dataframe.
This already enabled us to extract the specimen name and scan.
Since we performed multiple scans for each specimen, i.e.~a low resolution scan with large field of view for the whole head and one or two scans in high resolution focusing on the region of the oral and pharyngeal jaws it was necessary to be able to efficiently find the scan to be looked at in detail.
From the log files we extracted the scanning and reconstruction parameters of each performed scan for double-checking them and excluding (or necessarily correcting) any operator-error.
All relevant values for each scan were also saved into the aforementioned ataframe.
This allows for an easy collation of all the relevant values into a table (as shown in the \protect\hyperlink{supplementary-materials}{Supplementary Materials}) at the end of each run of the notebook.

After displaying several parameters of both data acquisition and reconstruction for ruling out any operator error we used \emph{Dask} {[}\protect\hyperlink{ref-hj1CnyWB}{22}{]} to efficiently access the tomographic data on disk (in the end amounting to a total of nearly a million single images).
On average, each of the tomographic datasets contains around 2700 slices, so the total size of the acquired data (1.5 TB) exceeds the RAM size available on an average high-end workstation.
The use of \emph{Dask} (and more specifically \emph{dask-image} {[}\protect\hyperlink{ref-wIfTDBqd}{23}/{]}) facilitated on-demand loading of the needed data directly from disk for each of the specimen to be analyzed in each step of the notebooks.

@@

To extract the oral jaw (OJ) and pharyngeal jaw (PJ) of the fish, we used \href{https://www.slicer.org}{\emph{3D Slicer}} (Version 4.11.20210226) {[}\protect\hyperlink{ref-OsCpHsJH}{24}{]} extended with the \emph{SlicerMorph} tools {[}\protect\hyperlink{ref-MwdN6kPV}{25}{]} which aim to help biologists work with 3D specimen data.
The reconstructed image stacks were loaded into \href{https://www.slicer.org/wiki/Documentation/Labs/ImageStacks}{\emph{ImageStacks}}, depending on their size we reduced the image resolution (i.e.~downscaled the images) for this first step.
The three-dimensional volume was rendered via \href{https://slicer.readthedocs.io/en/latest/user_guide/modules/volumerendering.html}{\emph{VTK GPU Ray Casting}}.
A custom-made volume property was used as an input to view the scans.
Using toggles in the volume rendering, we defined regions of interest (ROIs) for both the OJs and PJs in each specimen.
These ROIs were then extracted in their native resolution from the original dataset for further processing.
Using the gray value thresholding function in \emph{3D Slicer}'s \href{https://slicer.readthedocs.io/en/latest/user_guide/modules/segmenteditor.html}{\emph{Segment Editor}}, the teeth in both the oral and pharyngeal jaws were extracted.
We used the \emph{Scissor} and \emph{Island} tools of the \emph{Segment Editor} to isolate single regions.

Processed regions of interest were exported as \texttt{Nrrd} {[}\protect\hyperlink{ref-rn6XbmFW}{26}{]} files.
The three-dimensional visualizations of all regions of interest for each specimen were compiled into overview images (see Figure \ref{fig:104016} for an example from the compilation document).
In total, we compiled such overview images for 133 specimens with full head morphology, oral jaw and lower pharyngeal jaw profiles.

\begin{figure}
\hypertarget{fig:104016}{%
\centering
\includegraphics{images/104016.png}
\caption{Overview of data from sample 104016,~\emph{Enterochromis I cinctus}, from `Station E' in the transect of Mwanza Bay at the southern edge of Lake Victoria.
A map showing the detailed location of `Station E' is shown in Fig. 1 of {[}\protect\hyperlink{ref-JtHcU2O2}{27}{]}.
Panel A: Three-dimensional visualization of the head scan.
Panel B: Three-dimensional visualization of the oral jaw scan.
Panel C: Photograph of the specimen.
Panels D and E: Three-dimensional visualization of the pharyngeal jaw, dorsoventral and lateral view, respectively.}\label{fig:104016}
}
\end{figure}

@@

Current studies are using 3D geometric morphometrics to compare the morphological shape of these scanned cichlids using statistical analysis.
We used a homologous landmark scheme across one-half of the skull for higher density of shape information {[}\protect\hyperlink{ref-BDrCSu8p}{9},\protect\hyperlink{ref-15FafMZm5}{28}{]}, and landmarks were placed on each specimen using \emph{3D Slicer}.
To examine differences in shape across the species sampled, we performed a Generalized Procrustes Superimposition on the landmark data to remove the effects of location, size, and rotation from the analysis using the \emph{geomorph} package in \emph{R} (Version 4.2.1) with \emph{RStudio} (Version 2022.07.2+576) {[}\protect\hyperlink{ref-VPUOQA2P}{29}--\protect\hyperlink{ref-EwNL0C0Y}{31},\protect\hyperlink{ref-whEyWyQh}{32},\protect\hyperlink{ref-QfVloFVu}{33}{]}.
This process brings all specimens to a common origin, scales the landmarks to a unit centroid size, and rotates specimens to reduce distances between landmarks.
A principal components analysis was then performed in \emph{geomorph} on the superimposed landmark data to visualize the major axes of shape change across sampled species.
We then used phylogenetic information to identify instances of repeated evolution of trophic adaptations in these cichlids.

@@

Otoliths are structures made up of mostly calcium carbonate located in the head of fishes.
Due to their composition, they are easily distinguished in the X-ray images we acquired.
We devised an image processing method to automatically and robustly detect the location of the otoliths in the tomographic scans of the skulls of the cichlids.
Based on this detected location we were then able to efficiently extract a cropped region of interest from the original data which includes only the region of the otoliths.
The whole method is implemented in its own \href{https://github.com/habi/EAWAG/blob/master/ExtractOtoliths.ipynb}{\emph{Jupyter} notebook} (part of the aforementioned analysis repository {[}\protect\hyperlink{ref-1HteOscVd}{19}{]}).

@@

By detecting the largest components in the cropped copies of the datasets we can easily extract and visualize the otoliths in 3D, as shown in Figure \ref{fig:otolith3d}.
The extracted otoliths are thus prepared for further analysis and visualization.
The simple three-dimensional visualization is integrated in the aforementioned \emph{Jupyter} notebook through an integrated visualization library {[}\protect\hyperlink{ref-a9ZL41Of}{34}{]} and is also shown in the \protect\hyperlink{supplementary-materials}{Supplementary Materials}.

@@

The notebook for extracting the otoliths can be run in your browser without installing any additional software (via \emph{Binder} {[}\protect\hyperlink{ref-Q20Bxdsr}{35}{]}).
To do this, one starts the notebook by \href{https://mybinder.org/v2/gh/habi/eawag/HEAD?labpath=ExtractOtoliths.ipynb}{clicking a single button} in the \href{https://github.com/habi/EAWAG/blob/master/README.md}{README file} of the \href{https://github.com/habi/EAWAG}{project repository} {[}\protect\hyperlink{ref-1HteOscVd}{19}{]}.
This starts a computing environment in the cloud, downloads the tomographic data we acquired of \emph{one} specimen, and performs both the otolith extraction and visualization in your browser.

@@

The presented method offers an insight and algorithm on how to perform tomographic scans, preview and analyze micro-computer tomographic datasets of a large collection of fish.
The workflow is relying only on free and open-source software and can thus be used and verified independently by any interested reader.
All the \emph{Jupyter} notebook described herein is also freely available online {[}\protect\hyperlink{ref-1HteOscVd}{19}{]}.

@@

We thank Salome Mwaiko for taking care of the fish collection at Eawag and Mark Charran for helping to find suitable specimens to represent each species.
We are grateful to the \href{https//www.mic.unibe.ch/}{Microscopy Imaging Center} of the University of Bern for the infrastructural support.
We thank the \texttt{manubot} project {[}\protect\hyperlink{ref-YuJbg3zO}{36}{]} for helping us write this manuscript collaboratively.

@@

\leavevmode\vadjust pre{\hypertarget{ref-IKVBgaCE}{}}%
\CSLLeftMargin{7. }
\CSLRightInline{Rawson SD, Maksimcuka J, Withers PJ, Cartmell SH. X-ray computed tomography in life sciences. BMC Biol. 2020;18. doi:\href{https://doi.org/10.1186/s12915-020-0753-2}{10.1186/s12915-020-0753-2}}

\leavevmode\vadjust pre{\hypertarget{ref-8Pmc2mI8}{}}%
\CSLLeftMargin{8. }
\CSLRightInline{Messerli M, Aaldijk D, Haberthür D, Röss H, García-Poyatos C, Sande-Melón M, et al. Adaptation mechanism of the adult zebrafish respiratory organ to endurance training. Shelden EA, editor. PLoS ONE. 2020;15: e0228333. doi:\href{https://doi.org/10.1371/journal.pone.0228333}{10.1371/journal.pone.0228333}}

\leavevmode\vadjust pre{\hypertarget{ref-BDrCSu8p}{}}%
\CSLLeftMargin{9. }
\CSLRightInline{Ford KL, Peterson R, Bernt M, Albert JS. Convergence is Only Skin Deep: Craniofacial Evolution in Electric Fishes from South America and Africa (Apteronotidae and Mormyridae). Integrative Organismal Biology. 2022;4. doi:\href{https://doi.org/10.1093/iob/obac022}{10.1093/iob/obac022}}

\leavevmode\vadjust pre{\hypertarget{ref-3rxGCEvJ}{}}%
\CSLLeftMargin{10. }
\CSLRightInline{Available: \url{https://osf.io/ecmz4}}

\leavevmode\vadjust pre{\hypertarget{ref-pN1S9E5p}{}}%
\CSLLeftMargin{11. }
\CSLRightInline{Descamps E, Sochacka A, De Kegel B, Van Loo D, Van Hoorebeke L, Adriaens D. Soft tissue discrimination with contrast agents using micro-CT scanning. Belg J Zool. 2020;144. doi:\href{https://doi.org/10.26496/bjz.2014.63}{10.26496/bjz.2014.63}}

\leavevmode\vadjust pre{\hypertarget{ref-kkBa4qNV}{}}%
\CSLLeftMargin{12. }
\CSLRightInline{Babaei F, Hong TLC, Yeung K, Cheng SH, Lam YW. Contrast-Enhanced X-Ray Micro-Computed Tomography as a Versatile Method for Anatomical Studies of Adult Zebrafish. Zebrafish. 2016;13: 310--316. doi:\href{https://doi.org/10.1089/zeb.2016.1245}{10.1089/zeb.2016.1245}}

\leavevmode\vadjust pre{\hypertarget{ref-VCZPOv2f}{}}%
\CSLLeftMargin{13. }
\CSLRightInline{Available: \url{https://github.com/TomoGraphics/Hol3Drs/blob/master/STL/EAWAG.Fish.stl}}

\leavevmode\vadjust pre{\hypertarget{ref-115PPSuQp}{}}%
\CSLLeftMargin{14. }
\CSLRightInline{Haberthür D. TomoGraphics/Hol3Drs: A release. Zenodo; 2019. doi:\href{https://doi.org/10.5281/zenodo.2587555}{10.5281/zenodo.2587555}}

\leavevmode\vadjust pre{\hypertarget{ref-wsjdcTeC}{}}%
\CSLLeftMargin{15. }
\CSLRightInline{OpenSCAD. {[}cited 13 Jul 2023{]}. Available: \url{https://openscad.org}}

\leavevmode\vadjust pre{\hypertarget{ref-f7OzrzUv}{}}%
\CSLLeftMargin{16. }
\CSLRightInline{Available: \url{https://github.com/TomoGraphics/Hol3Drs/blob/master/EAWAG.Fish.scad}}

\leavevmode\vadjust pre{\hypertarget{ref-ZjgrPHTa}{}}%
\CSLLeftMargin{17. }
\CSLRightInline{Available: \url{https://github.com/habi/EAWAG/blob/master/rsync-fishes.sh}}

\leavevmode\vadjust pre{\hypertarget{ref-pQ6Wbz73}{}}%
\CSLLeftMargin{18. }
\CSLRightInline{Kluyver T, Ragan-Kelley B, Pérez F, Granger B, Bussonnier M, Frederic J, et al. Jupyter Notebooks -- a publishing format for reproducible computational workflows. In: Loizides F, Scmidt B, editors. IOS Press; 2016. pp. 87--90. doi:\href{https://doi.org/10.3233/978-1-61499-649-1-87}{10.3233/978-1-61499-649-1-87}}

\leavevmode\vadjust pre{\hypertarget{ref-1HteOscVd}{}}%
\CSLLeftMargin{19. }
\CSLRightInline{Haberthür D. habi/EAWAG: Jupyter notebooks for preparing tomographic scans of a collection of cichlids for analysis. Zenodo; 2022. doi:\href{https://doi.org/10.5281/zenodo.6798632}{10.5281/zenodo.6798632}}
